---
title: "Trends of Offerings and Purchases in Online Grocery Shopping"
output: 
  prettydoc::html_pretty:
    theme: leonids
    css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Overview

This project utilized data “The Instacart Online Grocery Shopping Dataset 2017” or the `instacart` dataset from package `p8105.datasets` to explore the trends of offerings and purchases in online grocery shopping. The project utilized R version 4.3.1 and R packages `dplyr`, `ggplot2`, `tidyverse`, and `p8105.datasets`. 

```{r load dataset & necessary R packages, message = FALSE, results = 'hide', echo = FALSE}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(p8105.datasets)
data("instacart")
head(instacart)
summary(instacart)
sum(is.na(instacart))
```

The **instacart** dataset has `r nrow(instacart)` observations and `r ncol(instacart)` variables, including: 

* `order_id`: order identifier, class `r class(pull(instacart, order_id))`.
* `product_id`: product identifier, class `r class(pull(instacart, product_id))`.
* `add_to_cart_order`: order in which each product was added to cart, class `r class(pull(instacart, add_to_cart_order))`.
* `reordered`: 1 if this prodcut has been ordered by this user in the past, 0 otherwise, class `r class(pull(instacart, reordered))`.
* `user_id`: customer identifier, class `r class(pull(instacart, user_id))`.
* `eval_set`: which evaluation set this order belongs in (Note that the data for use in this class is exclusively from the “train” eval_set), class `r class(pull(instacart, eval_set))`.
* `order_number`: the order sequence number for this user (1=first, n=nth), class `r class(pull(instacart, order_number))`.
* `order_dow`: the day of the week on which the order was placed, class `r class(pull(instacart, order_dow))`.
* `order_hour_of_day`: the hour of the day on which the order was placed, class `r class(pull(instacart, order_hour_of_day))`.
* `days_since_prior_order`: days since the last order, capped at 30, NA if order_number=1, class `r class(pull(instacart, days_since_prior_order))`.
* `product_name`: name of the product, class `r class(pull(instacart, product_name))`.
* `aisle_id`: aisle identifier, class `r class(pull(instacart, aisle_id))`.
* `department_id`: department identifier, class `r class(pull(instacart, department_id))`.
* `aisle`: the name of the aisle, class `r class(pull(instacart, aisle))`.
* `department`: the name of the department, class `r class(pull(instacart, department))`.

## Visualizations and EDA

#### Table for Number of Orders, by Day of Week
```{r Explore key variables, message = FALSE, echo = FALSE}
#Summary of unique users
users_sum <- instacart |> 
  group_by(user_id) |> 
  summarize(n_obs = n())

#Summary of unique orders
orders_sum <- instacart |> 
  group_by(order_id) |> 
  summarize(n_obs = n())

#Summary of aisles
aisle_sum <- instacart |> 
  group_by(aisle) |> 
  summarize(n_obs = n())

#Summary of departments
department_sum <- instacart |> 
  group_by(department) |> 
  summarize(n_obs = n())

#Summary of sales by the day of the week on which the order was placed
sum_dow <- instacart |>
  group_by(order_dow) |>
  summarize(n_obs = n())
sum_dow |> knitr::kable()
options(scipen = 999)
```

There are `r nrow(users_sum)` unique users and `r nrow(orders_sum)` unique orders. On average, there are `r round(mean(pull(sum_dow,n_obs)), 2)` items sold per day. There are `r length(unique(pull(aisle_sum, aisle)))` aisles and `r length(unique(pull(department_sum, department)))` departments. There are `r length(unique(pull(instacart, aisle)))` aisles, and most items are ordered from aisle *`r filter(aisle_sum, n_obs == max(pull(aisle_sum,n_obs))) |> select(aisle)`*. 

#### Department Barplot
```{r Department Barplot, echo = FALSE, message = FALSE}
#departments, in barplot
instacart |> ggplot(aes(y = department)) +
  geom_bar(color = "blue", fill = "lightblue") +
  labs(title ="Department Barplot",
       y ='Department',
       x ='Frequency'
      ) + 
  theme(axis.text.y=element_text(size=10))
```

#### Aisles Barplot, colored by departments
```{r Aisles Barplot, echo = FALSE, message = FALSE}
#aisles barplot, colored by departments
instacart |> 
  ggplot(aes(x = aisle_id, color = department, fill = department)) + 
  geom_bar(position = "dodge") + 
  labs(title = "Aisles Histogram",
       x = "Unique Aisle ID",
       y = "Number of Items (Count)"
      ) + 
  theme(legend.position = "top")
```

#### Number of items ordered in each aisle, for aisles with more than 10000 items ordered. 
```{r Q1- Making the aisle plot, echo = FALSE}
#aisles, in barplot
aisle_for_plot <- instacart |>
  group_by(aisle) |>
  summarize(n_obs = n()) |>
  filter(n_obs > 10000)

aisle_for_plot |>
  ggplot(aes(y = aisle, x = n_obs)) +
  geom_point() +
  labs(title = "Aisle Scatterplot",
       x ='Number of Items Purchased',
       y ='Aisle'
      ) + 
  theme(axis.text.y=element_text(size=8))
```

There are `r length(unique(pull(aisle_for_plot, aisle)))` aisles with more than 10,000 items ordered. Most aisles have less than 40,000 items ordered. There are 5 aisles that have more than 40,000 items sold. They are: `r pull(filter(aisle_for_plot, n_obs > 40000), aisle)`.
  
#### Three most popular items in aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”
```{r Making the table showing three most popular items, message = FALSE, echo = FALSE}
aisle_for_table <- instacart |>
  filter(aisle == "baking ingredients" | aisle == "dog food care" | aisle == "packaged vegetables fruits") |>
  group_by(aisle, product_name) |>
  summarize(n_obs = n()) |>
  filter(n_obs %in% tail(sort(n_obs), 3)) |>
  arrange(desc(n_obs))
aisle_for_table |> knitr::kable()
```

Among the three aisles, `r aisle_for_table[which.max(pull(aisle_for_table, n_obs)), 2]` has the most sales of `r max(pull(aisle_for_table, n_obs))`.

#### Mean hour of day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week
```{r Making the table for mean hour of day, message = FALSE, echo = FALSE}
aisle_for_table_2 <- instacart |>
  filter(product_name == "Pink Lady Apples" | product_name == "Coffee Ice Cream") |>
  group_by(order_dow) |>
  mutate(mean_hour = mean(order_hour_of_day, na.rm = TRUE)) |>
  distinct(order_dow, mean_hour) |>
  arrange(order_dow) |>
  mutate(order_dow = recode(order_dow, "0" = "Sunday", "1" = "Monday", "2" = "Tuesday", 
                            "3" = "Wednesday", "4" = "Thursday", "5" = "Friday", "6" = "Saturday")) |>
  pivot_wider(names_from = "order_dow", values_from = "mean_hour") 
aisle_for_table_2 |> knitr::kable()
```
  
`r colnames(aisle_for_table_2[max.col(aisle_for_table_2,ties.method="first")])` has the latest mean hour of day at which Pink Lady Apples and Coffee Ice Cream are ordered. 