---
title: "p8105_hw3_xl3214"
author: "Xuan Lu"
date: "2023-10-11"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

```{r load dataset & necessary R packages, message = FALSE, results = 'hide'}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(p8105.datasets)
data("instacart")
head(instacart)
summary(instacart)
sum(is.na(instacart))
```

The **instacart** dataset has `r nrow(instacart)` observations and `r ncol(instacart)` variables, including: 

* `order_id`: order identifier, class `r class(pull(instacart, order_id))`.

* `product_id`: product identifier, class `r class(pull(instacart, product_id))`.

* `add_to_cart_order`: order in which each product was added to cart, class `r class(pull(instacart, add_to_cart_order))`.

* `reordered`: 1 if this prodcut has been ordered by this user in the past, 0 otherwise, class `r class(pull(instacart, reordered))`.

* `user_id`: customer identifier, class `r class(pull(instacart, user_id))`.

* `eval_set`: which evaluation set this order belongs in (Note that the data for use in this class is exclusively from the “train” eval_set), class `r class(pull(instacart, eval_set))`.

* `order_number`: the order sequence number for this user (1=first, n=nth), class `r class(pull(instacart, order_number))`.

* `order_dow`: the day of the week on which the order was placed, class `r class(pull(instacart, order_dow))`.

* `order_hour_of_day`: the hour of the day on which the order was placed, class `r class(pull(instacart, order_hour_of_day))`.

* `days_since_prior_order`: days since the last order, capped at 30, NA if order_number=1, class `r class(pull(instacart, days_since_prior_order))`.

* `product_name`: name of the product, class `r class(pull(instacart, product_name))`.

* `aisle_id`: aisle identifier, class `r class(pull(instacart, aisle_id))`.

* `department_id`: department identifier, class `r class(pull(instacart, department_id))`.

* `aisle`: the name of the aisle, class `r class(pull(instacart, aisle))`.

* `department`: the name of the department, class `r class(pull(instacart, department))`.

```{r Explore key variables, message = FALSE, echo = FALSE}
#Summary of unique users
users_sum <- instacart |> 
  group_by(user_id) |> 
  summarize(n_obs = n())

#Summary of unique orders
orders_sum <- instacart |> 
  group_by(order_id) |> 
  summarize(n_obs = n())

#Summary of aisles
aisle_sum <- instacart |> 
  group_by(aisle) |> 
  summarize(n_obs = n())

#Summary of departments
department_sum <- instacart |> 
  group_by(department) |> 
  summarize(n_obs = n())

#Summary of sales by the day of the week on which the order was placed
sum_dow <- instacart |>
  group_by(order_dow) |>
  summarize(n_obs = n())
sum_dow
```

  * There are `r nrow(users_sum)` unique users and `r nrow(orders_sum)` unique orders. On average, there are `r mean(pull(sum_dow,n_obs))` items sold per day. There are `r length(unique(pull(aisle_sum, aisle)))` aisles and `r length(unique(pull(department_sum, department)))` departments. Below are some graphs to help visualize the dataset key variables.

```{r Q1- Some basic visualizations, echo = FALSE, message = FALSE}
#number of unique orders and distribution of total items purchased per order, colored by unique users
instacart |> 
  group_by(order_id) |> 
  summarize(n_obs = n()) |> 
  ggplot(aes(x = n_obs)) + 
  geom_histogram(color = "blue", fill = "lightblue") + 
  labs(
    title = "Orders Histogram",
    x = "Unique Order ID",
    y = "Number of Items (Count)"
  )

#departments, in barplot
instacart |> ggplot(aes(y = department)) +
  geom_bar(color = "blue", fill = "lightblue") +
  labs(title ="Department Barplot",
       y ='Department',
       x ='Frequency'
      ) + 
  theme(axis.text.y=element_text(size=10))

#aisles, in barplot
instacart |> ggplot(aes(x = aisle_id)) +
  geom_bar(color = "blue", fill = "lightblue") +
  labs(title = "Aisle Barplot",
       x ='Unique Aisle ID',
       y ='Frequency'
      )

#aisles barplot, colored by departments
instacart |> 
  ggplot(aes(x = aisle_id, color = department, fill = department)) + 
  geom_bar(position = "dodge") + 
  labs(title = "Aisles Histogram",
       x = "Unique Aisle ID",
       y = "Number of Items (Count)"
      ) + 
  theme(legend.position = "top")
```

1. How many aisles are there, and which aisles are the most items ordered from?

  * Answer: There are `r length(unique(pull(instacart, aisle)))` aisles, and most items are ordered from aisle *`r filter(aisle_sum, n_obs == max(pull(aisle_sum,n_obs))) |> select(aisle)`*.

2. Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.

```{r Q1- Making the aisle plot, echo = FALSE}
#aisles, in barplot
aisle_for_plot <- instacart |>
  group_by(aisle) |>
  summarize(n_obs = n()) |>
  filter(n_obs > 10000)

aisle_for_plot |>
  ggplot(aes(y = aisle, x = n_obs)) +
  geom_point() +
  labs(title = "Aisle Scatterplot",
       x ='Number of Items Purchased',
       y ='Aisle'
      ) + 
  theme(axis.text.y=element_text(size=8))
```

  * There are `r length(unique(pull(aisle_for_plot, aisle)))` aisles with more than 10,000 items ordered. 
  
  * Most aisles have less than 40,000 items ordered. There are 5 aisles that have more than 40,000 items sold. They are: `r pull(filter(aisle_for_plot, n_obs > 40000), aisle)`.
  

3. Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.

```{r Making the table showing three most popular items, message = FALSE, echo = FALSE}
aisle_for_table <- instacart |>
  filter(aisle == "baking ingredients" | aisle == "dog food care" | aisle == "packaged vegetables fruits") |>
  group_by(aisle, product_name) |>
  summarize(n_obs = n()) |>
  filter(n_obs %in% tail(sort(n_obs), 3)) |>
  arrange(desc(n_obs))
aisle_for_table
```

  * For aisle *packaged vegetables fruits*, products `r pull(filter(aisle_for_table, aisle == "packaged vegetables fruits"), product_name)` are the three most popular products, each correspond to `r pull(filter(aisle_for_table, aisle == "packaged vegetables fruits"), n_obs)` in sales.

  * For aisle *baking ingredients*, products `r pull(filter(aisle_for_table, aisle == "baking ingredients"), product_name)` are the three most popular products, each correspond to `r pull(filter(aisle_for_table, aisle == "baking ingredients"), n_obs)` in sales.
  
  * For aisle *dog food care*, products `r pull(filter(aisle_for_table, aisle == "dog food care"), product_name)` are the three most popular products, each correspond to `r pull(filter(aisle_for_table, aisle == "dog food care"), n_obs)` in sales.
  
  * Among the three aisles, `r aisle_for_table[which.max(pull(aisle_for_table, n_obs)), 2]` has the most sales of `r max(pull(aisle_for_table, n_obs))`.


4. Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).

```{r Making the table for mean hour of day, message = FALSE, echo = FALSE}
aisle_for_table_2 <- instacart |>
  filter(product_name == "Pink Lady Apples" | product_name == "Coffee Ice Cream") |>
  group_by(order_dow) |>
  mutate(mean_hour = mean(order_hour_of_day, na.rm = TRUE)) |>
  distinct(order_dow, mean_hour) |>
  arrange(order_dow) |>
  mutate(order_dow = recode(order_dow, "0" = "Sun", "1" = "Mon", "2" = "Tue", 
                            "3" = "Wed", "4" = "Thurs", "5" = "Fri", "6" = "Sat")) |>
  pivot_wider(names_from = "order_dow", values_from = "mean_hour")
aisle_for_table_2
```
  
  * `r colnames(aisle_for_table_2[max.col(aisle_for_table_2,ties.method="first")])` has the latest mean hour of day at which Pink Lady Apples and Coffee Ice Cream are ordered. 


## Question 2
```{r load dataset, message = FALSE, results = 'hide'}
data("brfss_smart2010")
sum(is.na(brfss_smart2010))
```

```{r data cleaning, results = 'hide', message = FALSE}
#format the data to use appropriate variable names
#focus on the “Overall Health” topic, include only responses from “Excellent” to “Poor”
#organize responses as a factor taking levels ordered from “Poor” to “Excellent”
brfss <- brfss_smart2010 |>
  janitor::clean_names() |> 
  filter(topic == "Overall Health" & response %in% c("Excellent", "Very good", "Good", "Fair", "Poor")) |> 
  mutate(response = factor(response, 
                           levels=c("Poor", "Fair", "Good", "Very good", "Excellent"), ordered=TRUE)) |>
  rename(resp_id = respid, location_abbr = locationabbr, location_desc = locationdesc)
head(brfss)
sum(is.na(brfss))
unique(pull(brfss, topic))
unique(pull(brfss, response))
nrow(!is.na(brfss))
```

The BRFSS dataset has `r nrow(brfss)` observations and `r ncol(brfss)` variables. Variables include: `r colnames(brfss)`.

1. In 2002, which states were observed at 7 or more locations? What about in 2010?

```{r results = 'hide', message = FALSE, echo = FALSE}
brfss_q1 <- brfss |> 
  group_by(year) |> 
  select(year, location_abbr, location_desc) |>
  distinct(year, location_abbr, location_desc) |>
  group_by(year, location_abbr) |>
  summarise(n_obs = n())
sum(is.na(brfss_q1))
```

  * In 2002, `r pull(filter(brfss_q1, year == "2002" & n_obs > 7), location_abbr)` states were observed at 7 or more locations. 
  
  * In 2010, `r pull(filter(brfss_q1, year == "2010" & n_obs > 7), location_abbr)` states were observed at 7 or more locations. 

2. Construct a dataset that is limited to Excellent responses, and contains, year, state, and a variable that averages the data_value across locations within a state. Make a “spaghetti” plot of this average value over time within a state (that is, make a plot showing a line for each state across years – the geom_line geometry and group aesthetic will help).

```{r Q2- Spaghetti Plot, echo = FALSE, warning = FALSE}
brfss_q2 <- brfss |> 
  filter(response == "Excellent") |>
  group_by(year, location_abbr) |> 
  mutate(data_value_state_avg = mean(data_value)) |>
  select(year, location_abbr, data_value_state_avg) |>
  distinct(year, location_abbr, data_value_state_avg)
brfss_q2
check_na <- which(is.na(brfss_q2))
brfss_q2[check_na,]

ggplot(brfss_q2, aes(x = year, y = data_value_state_avg, color = location_abbr)) + 
  geom_line() +
  labs(title = "Spaghetti Plot: Yearly Average Data Value, by State",
       x = "Year",
       y = "Average Data Value"
      )
```

  * The average data values across locations within a state fluctuates across years, but overall remains fairly stable. There is no distinguishable trend in the average data values across years. 
  
  * Among all data available, `r brfss_q2[which.max(pull(brfss_q2,data_value_state_avg)), 2]` in `r brfss_q2[which.max(pull(brfss_q2,data_value_state_avg)), 1]` has the highest average data value, `r brfss_q2[which.min(pull(brfss_q2,data_value_state_avg)), 2]` in `r brfss_q2[which.min(pull(brfss_q2,data_value_state_avg)), 1]` has the lowest average data value.

3. Make a two-panel plot showing, for the years 2006, and 2010, distribution of data_value for responses (“Poor” to “Excellent”) among locations in NY State.

```{r Q2- Two-panel plot, echo = FALSE, warning = FALSE}
brfss_q3 <- brfss |>
  filter(location_abbr == "NY" & year %in% c("2006", "2010")) |>
  select(year, location_desc, response, data_value)

#Graph 1: without coloring by county
ggplot(brfss_q3, aes(x = response, y = data_value)) + 
  geom_boxplot(outlier.colour="black", outlier.shape=16,
             outlier.size=2, notch=FALSE) + 
  facet_grid(. ~ year) +
  labs(
    title = "Box Plot: Data Value for Responses in NY State, 2006 vs. 2010",
       x = "Response",
       y = "Data Value"
      ) + 
  theme(
    title = element_text(size=10),
    axis.text.x = element_text(size=7))

#Graph 2: with coloring by county
ggplot(brfss_q3, aes(x = response, y = data_value, color = location_desc)) + 
  geom_point() + 
  facet_grid(. ~ year) +
  labs(
    title = "Scatter Plot: Data Value for Responses among Locations in NY State, 2006 vs. 2010",
       x = "Response",
       y = "Data Value"
      ) + 
  theme(
    title = element_text(size=10),
    axis.text.x = element_text(size=7))
```

  * Based on the plots, we can see that *Poor* responses correspond to the lowest data values across all locations in NY State for both 2006 and 2010, and as the order of responses increases towards *Very good*, the data values also increase. However, for responses of *Excellent*, we see a decreased data value distribution compared to the previous *Good* and *Very good*. 
  
  * We can see that in 2006, responses *Good* and *Very good* have comparable data value distributions, although the median for *Very good* is slightly higher. When it comes to 2010, the data value for responses *Very good* is apparently higher than that of *Good*. Responses *Fair* and *Poor* in 2010 have wider spread compared to that of 2006.

## Question 3

1. Load, tidy, merge, and otherwise organize the data sets. 

Your final dataset should include all originally observed variables; exclude participants less than 21 years of age, and those with missing demographic data; and encode data with reasonable variable classes (i.e. not numeric, and using factors with the ordering of tables and plots in mind).

```{r Q3: Read-in datasets and cleaning, message = FALSE, results = 'hide'}
#Checking variable descriptions (first few rows)
covar_desc <- read_csv(file = "./nhanes_covar.csv", col_names = FALSE)
covar_desc <- covar_desc[1:4,]
covar_desc

covar <- read_csv(file = "./nhanes_covar.csv", skip = 4, col_names = TRUE) |> 
  janitor::clean_names() |> 
  mutate(sex = recode(sex, "1" = "male", "2" = "female")) |>
  mutate(education = recode(education, "1" = "Less than high school", "2" = "High school equivalent", "3" = "More than high school")) |>
  filter(age < 21)

#Remove NAs
covar <- na.omit(covar)
```

Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category. Comment on these items.

Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot.

Accelerometer data allows the inspection activity over the course of the day. Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences.





