---
title: "p8105_hw3_xl3214"
author: "Xuan Lu"
date: "2023-10-11"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

```{r load dataset & necessary R packages, message = FALSE}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(p8105.datasets)
data("instacart")
```

The **instacart** dataset has `r nrow(instacart)` observations and `r ncol(instacart)` variables, including: 

* `order_id`: order identifier, class `r class(pull(instacart, order_id))`.

* `product_id`: product identifier, class `r class(pull(instacart, product_id))`.

* `add_to_cart_order`: order in which each product was added to cart, class `r class(pull(instacart, add_to_cart_order))`.

* `reordered`: 1 if this prodcut has been ordered by this user in the past, 0 otherwise, class `r class(pull(instacart, reordered))`.

* `user_id`: customer identifier, class `r class(pull(instacart, user_id))`.

* `eval_set`: which evaluation set this order belongs in (Note that the data for use in this class is exclusively from the “train” eval_set), class `r class(pull(instacart, eval_set))`.

* `order_number`: the order sequence number for this user (1=first, n=nth), class `r class(pull(instacart, order_number))`.

* `order_dow`: the day of the week on which the order was placed, class `r class(pull(instacart, order_dow))`.

* `order_hour_of_day`: the hour of the day on which the order was placed, class `r class(pull(instacart, order_hour_of_day))`.

* `days_since_prior_order`: days since the last order, capped at 30, NA if order_number=1, class `r class(pull(instacart, days_since_prior_order))`.

* `product_name`: name of the product, class `r class(pull(instacart, product_name))`.

* `aisle_id`: aisle identifier, class `r class(pull(instacart, aisle_id))`.

* `department_id`: department identifier, class `r class(pull(instacart, department_id))`.

* `aisle`: the name of the aisle, class `r class(pull(instacart, aisle))`.

* `department`: the name of the department, class `r class(pull(instacart, department))`.

```{r Explore key variables, message = FALSE, echo = FALSE}
#Summary of unique users
users_sum <- instacart |> 
  group_by(user_id) |> 
  summarize(n_obs = n())

#Summary of unique orders
orders_sum <- instacart |> 
  group_by(order_id) |> 
  summarize(n_obs = n())

#Summary of aisles
aisle_sum <- instacart |> 
  group_by(aisle) |> 
  summarize(n_obs = n())

#Summary of departments
department_sum <- instacart |> 
  group_by(department) |> 
  summarize(n_obs = n())

#Summary of sales by the day of the week on which the order was placed
sum_dow <- instacart |>
  group_by(order_dow) |>
  summarize(n_obs = n())
sum_dow
```

  * There are `r nrow(users_sum)` unique users and `r nrow(orders_sum)` unique orders. On average, there are `r mean(sum_dow$n_obs)` items sold per day. There are `r length(unique(pull(aisle_sum, aisle)))` aisles and `r length(unique(pull(department_sum, department)))` departments. Below are some graphs to help visualize the dataset key variables.

```{r Some basic visualizations, echo = FALSE, message = FALSE}
#number of unique orders and distribution of total items purchased per order, colored by unique users
instacart |> 
  group_by(order_id) |> 
  summarize(n_obs = n()) |> 
  ggplot(aes(x = n_obs)) + 
  geom_histogram(color = "blue", fill = "lightblue") + 
  labs(
    title = "Orders Histogram",
    x = "Unique Order ID",
    y = "Number of Items (Count)"
  )

#departments, in barplot
instacart |> ggplot(aes(y = department)) +
  geom_bar(color = "blue", fill = "lightblue") +
  labs(title ="Department Barplot",
       y ='Department',
       x ='Frequency'
      ) + 
  theme(axis.text.y=element_text(size=10))

#aisles, in barplot
instacart |> ggplot(aes(x = aisle_id)) +
  geom_bar(color = "blue", fill = "lightblue") +
  labs(title = "Aisle Barplot",
       x ='Unique Aisle ID',
       y ='Frequency'
      )

#aisles barplot, colored by departments
instacart |> 
  ggplot(aes(x = aisle_id, color = department, fill = department)) + 
  geom_bar(position = "dodge") + 
  labs(title = "Aisles Histogram",
       x = "Unique Aisle ID",
       y = "Number of Items (Count)"
      ) + 
  theme(legend.position = "top")
```

1. How many aisles are there, and which aisles are the most items ordered from?

  * Answer: There are `r length(unique(pull(instacart, aisle)))` aisles, and most items are ordered from aisle *`r filter(aisle_sum, n_obs == max(aisle_sum$n_obs)) |> select(aisle)`*.

2. Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.

```{r Making the plot, echo = FALSE}
#aisles, in barplot
aisle_for_plot <- instacart |>
  group_by(aisle) |>
  summarize(n_obs = n()) |>
  filter(n_obs > 10000)

aisle_for_plot |>
  ggplot(aes(y = aisle, x = n_obs)) +
  geom_point() +
  labs(title = "Aisle Scatterplot",
       x ='Number of Items Purchased',
       y ='Aisle'
      ) + 
  theme(axis.text.y=element_text(size=8))
```

  * There are `r length(unique(pull(aisle_for_plot, aisle)))` aisles with more than 10,000 items ordered. 
  
  * Most aisles have less than 40,000 items ordered. There are 5 aisles that have more than 40,000 items sold. They are: `r pull(filter(aisle_for_plot, n_obs > 40000), aisle)`.
  

3. Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.

```{r Making the table showing three most popular items, message = FALSE, echo = FALSE}
aisle_for_table <- instacart |>
  filter(aisle == "baking ingredients" | aisle == "dog food care" | aisle == "packaged vegetables fruits") |>
  group_by(aisle, product_name) |>
  summarize(n_obs = n()) |>
  filter(n_obs %in% tail(sort(n_obs), 3)) |>
  arrange(desc(n_obs))
aisle_for_table
```

  * For aisle *packaged vegetables fruits*, products `r pull(filter(aisle_for_table, aisle == "packaged vegetables fruits"), product_name)` are the three most popular products, each correspond to `r pull(filter(aisle_for_table, aisle == "packaged vegetables fruits"), n_obs)` in sales.

  * For aisle *baking ingredients*, products `r pull(filter(aisle_for_table, aisle == "baking ingredients"), product_name)` are the three most popular products, each correspond to `r pull(filter(aisle_for_table, aisle == "baking ingredients"), n_obs)` in sales.
  
  * For aisle *dog food care*, products `r pull(filter(aisle_for_table, aisle == "dog food care"), product_name)` are the three most popular products, each correspond to `r pull(filter(aisle_for_table, aisle == "dog food care"), n_obs)` in sales.
  
  * Among the three aisles, `r aisle_for_table[which.max(pull(aisle_for_table, n_obs)), 2]` has the most sales of `r max(pull(aisle_for_table, n_obs))`.


4. Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).

```{r Making the table for mean hour of day, message = FALSE, echo = FALSE}
aisle_for_table_2 <- instacart |>
  filter(product_name == "Pink Lady Apples" | product_name == "Coffee Ice Cream") |>
  group_by(order_dow) |>
  mutate(mean_hour = mean(order_hour_of_day, na.rm = TRUE)) |>
  distinct(order_dow, mean_hour) |>
  arrange(order_dow) |>
  mutate(order_dow = recode(order_dow, "0" = "Sun", "1" = "Mon", "2" = "Tue", 
                            "3" = "Wed", "4" = "Thurs", "5" = "Fri", "6" = "Sat")) |>
  pivot_wider(names_from = "order_dow", values_from = "mean_hour")
aisle_for_table_2
```
  
  * `r colnames(aisle_for_table_2[max.col(aisle_for_table_2,ties.method="first")])` has the latest mean hour of day at which Pink Lady Apples and Coffee Ice Cream are ordered. 


## Question 2
```{r load dataset, message = FALSE}
data("brfss_smart2010")
```

